\documentclass{article}
\usepackage{amsmath}
\usepackage{tikz}
\usepackage[margin=1in]{geometry}



\usetikzlibrary{calc}


\setlength{\parskip}{4pt}
\title{Cars 4 -- Lightnight McQuation}
\author{Rhitt C}
\date{Feb 2026}


\begin{document}
\maketitle

\section*{Abstract}
Lightning McQueen, old and rusty, decides to officially retire from the racing scene, leaving Cruz to carry on his legacy.

In search of a new career, he thinks back to all the crazy stunts he'd pulled over the years and realises he has no idea how they were even possible physically.
So he has a crazy idea: why not become a physicist and research stunt mechanics?

Who knows, maybe a strong understanding will lay the foundations for a future NASCAR training business $\dots$

Note to self: learn simulink and experiment further with simulations

\section{The Basics of Friction -- Walking}

\section{First Paradigm -- Skidding and Drifting}

\section{Second Paradigm -- Banked Tracks}

\subsection{Deriving a general formula for Friction and Normal forces}
The first constraint is that the car travels in a purely horizontal plane 
  i.e. there is no vertical movement and so net vertical force on the car is $\vec{0}$.

For ease of calculation, we deal with both friction down the slope (\textbf{Case 1}) and up the slope (\textbf{Case 2}) at once using ($\pm$) and ($\mp$).

Throughout the calculations, the sign on top (i.e. ($+$) in ($\pm$) and ($-$) in ($\mp$)) is for Case 1 and the sign on bottom for Case 2.

\begin{figure}[h]
\centering
\begin{tikzpicture}
    \pgfmathsetmacro{\radius}{2}
    \pgfmathsetmacro{\g}{9.81}
    \pgfmathsetmacro{\mass}{0.1}
    \pgfmathsetmacro{\incline}{atan(3/4)}
    \pgfmathsetmacro{\fric}{0.3}
    \pgfmathsetmacro{\speed}{sqrt(\radius * \g * (  (tan(\incline)+\fric)/(1-\fric*tan(\incline))  ))}

    \draw (0, 0) -- (4, 0) -- (4, 3);
    \draw (4, 3) -- (0, 0) node[midway, above, sloped, draw, fill=blue!20, minimum size=10pt] (c) {};


    \draw[dashed, ->] (6, 1) -- (6, 1.5) node[above, font=\tiny] {$(+)\text{ve }y$};
    \draw[dashed, ->] (6, 1) -- (5.5, 1) node[left, font=\tiny] {$(+)\text{ve }x$};


    
    \pgfmathsetmacro{\Fg}{\mass * \g}
    \fill[cyan] (c.center) circle (2pt);
    \draw[cyan, thick, ->] (c.center) -- ++(0, -\Fg)
        node[below] {$\vec{F}_g$};

    \pgfmathsetmacro{\FN}{\mass * ((\speed^2 * sin(\incline) / \radius) + (\g * cos(\incline)))}
    \fill[magenta] (c.south) circle (2pt);
    \draw[magenta, thick, ->] 
        (c.south) -- ++({\FN*cos(\incline+90)},{\FN*sin(\incline+90)})
        node[left] {$\vec F_N$};

    \pgfmathsetmacro{\Ff}{\mass * ((\speed^2 * cos(\incline) / \radius) - (\g * sin(\incline)))}
    \fill[magenta] (c.south) circle (2pt);
    \draw[magenta, thick, ->] 
        (c.south) -- ++({\Ff*cos(\incline+180)},{\Ff*sin(\incline+180)})
        node[left] {$\vec F_f$};
\end{tikzpicture}
\caption{Toy Car on banked track with $m=0.1$, $r=2$, $\theta=\arctan\left(\frac{3}{4}\right)$, $\mu=0.3$, $g=9.81$ SI units}
\end{figure}

\pagebreak


Setting the upwards direction as positive, we have the following:
\begin{align*}
    \sum \vec{F}_y &= \vec{0} \\
    \vec{0}&=\vec{F_g} + \vec{F}_{fy} + \vec{F}_{Ny} \\
    \vec{0}&=-mg \mp |F_f|\sin\theta + |F_N|\cos\theta \tag{E1}\\
    (\text{E1})\cdot \cos\theta\implies\qquad\qquad 
        \vec{0} &=-mg\cos\theta \mp |F_f|\sin\theta\cos\theta + |F_N|\cos^2\theta \tag{E2}\\
\end{align*}


The second constraint is that the car undergoes uniform circular motion.

Setting the direction in towards the center of the velodrome to be positive, we must have positive centripetal force and hence:
\begin{align*}
    \sum \vec{F}_x &= +\; m\frac{v^2}{r} \\
    \vec{F}_{fx} + \vec{F}_{Nx}&=m\frac{v^2}{r} \\
    \vec{0}&= m\frac{v^2}{r} - \vec{F}_{fx} - \vec{F}_{Nx} \\
    \vec{0}&= m\frac{v^2}{r} \mp |F_f|\cos\theta - |F_N|\sin\theta \tag{E3}\\
    (\text{E3})\cdot \sin\theta\implies\qquad\qquad 
        \vec{0}&= m\frac{v^2}{r}\sin\theta \mp |F_f|\sin\theta\cos\theta - |F_N|\sin^2\theta \tag{E4}\\
\end{align*}


Now we combine to obtain an equation involving $F_f$ but not $F_N$

\begin{align*}
    \text{(E4)} - \text{(E2)} \implies \qquad
        \vec{0}-\vec{0}  &=  m\frac{v^2}{r}\sin\theta + mg\cos\theta + (\vec{0}) + |F_N|\cdot (-\sin^2\theta - \cos^2\theta) \\ 
        \vec{0} &= m\frac{v^2}{r}\sin\theta + mg\cos\theta + |F_N|\cdot(-1) \\ 
        |F_N| &= m \left(\frac{v^2}{r} \sin\theta + g\cos\theta \right) \tag{E5}
\end{align*}


This gives us the normal force! Now, plugging into (E1), we find the friction force

\begin{align*}
    \text{(E5)}\rightarrow\text{(E1)}\implies\quad\quad
        \vec{0}&=-mg \mp |F_f|\sin\theta + m\left(g\cos\theta + \frac{v^2}{r}\sin\theta\right)\cos\theta \\
        \pm |F_f|\sin\theta  &= m\left(\frac{v^2}{r}\sin\theta\cos\theta + g (-1+\cos^2 \theta)\right) \\
                             &= m\left(\frac{v^2}{r}\sin\theta\cos\theta + g (-\sin^2\theta)\right) \\
        |F_f| &= \pm m\left(\frac{v^2}{r} \cos\theta - g\sin\theta\right)
\end{align*}


Summarising our results thus far, we have,

\begin{center}
\boxed{\vec{F_N} = m\left(\frac{v^2}{r}\sin\theta + g\cos\theta\right)\text{ up perpendicular to slope}}
\end{center}

and splitting the answer for friction into the two cases:

\begin{center}
\boxed{\vec{F_f} = m\left( \frac{v^2}{r} \cos\theta - g\sin\theta\right)\text{ down the slope }\qquad\text{or}\qquad \vec{F}_f = m\left( g\sin\theta - \frac{v^2}{r} \cos\theta\right)\text{ up the slope}}
\end{center}

\subsection{Deriving Range of Speeds}

We now wish to find the maximum value of $|v|$ that meets our constraints for horizontal circular motion. This occurs when $F_f$ is at its maximum down the slope (i.e. ($\pm$) replaced with ($+$)), as this means $F_f$ has a maximal centre-seeking component, and so does $F_N$ as it increases to balance out both friction and gravity. Thus, they both contribute as much as possible to the centripetal force. Hence, we have:

\begin{align*}
|F_f|_{(+)} &\le \mu |F_N|\\
m\left(\frac{v^2}{r}\cos\theta - g \sin\theta\right)
&\le \mu m \left(\frac{v^2}{r}\sin\theta + g \cos\theta\right)\\
\frac{v^2}{r}\cos\theta - g \sin\theta
&\le \mu  \left(\frac{v^2}{r}\sin\theta + g \cos\theta\right)\\
\frac{v^2}{r}\left(\cos\theta - \mu\sin\theta\right)
&\le g (\sin\theta + \mu\cos\theta)\\
\frac{v^2}{r}&\le g\frac{\sin\theta + \mu\cos\theta}{\cos\theta - \mu\sin\theta}\\
|v|&\le \sqrt{rg\left(\frac{\sin\theta + \mu\cos\theta}{\cos\theta - \mu\sin\theta}\right)}
\end{align*}

Similarly, to minimise $|v|$, we want $F_f$ to be maximally directed up the slope (i.e. replace ($\pm$) with ($-$)). 
The primary reason is that this leads to a maximal component away from the center of the velodrome. The vertical component also ``lessens the burden'' of $F_N$ in opposing the constant $F_g$ downwards, and thereby weakens $F_N$ alongside its center-seeking component to a min.
Both consequences minimise centripetal force and thus speed as required.

Thus,

\begin{align*}
|F_f|_{(-)} &\le -\mu |F_N|\\
-m\left(\frac{v^2}{r}\cos\theta - g \sin\theta\right)
    &\le \mu m\left(\frac{v^2}{r}\sin\theta + g \cos\theta\right)\\
\frac{v^2}{r} \left(\cos\theta + \mu\sin\theta\right)
&\ge  g (\sin\theta - \mu\cos\theta)\\
|v| &\ge
\sqrt{rg\left(\frac{\sin\theta - \mu\cos\theta}{\cos\theta + \mu\sin\theta}\right)}
\end{align*}

Therefore, as static friction can take any magnitude up to the threshold necessary to maintain our first constraint,
we have the full range of speeds for which static friction and the normal force can work to ensure there is horizontal circular motion:

\begin{center}\boxed{\sqrt{rg\left(\frac{\sin\theta + \mu\cos\theta}{\cos\theta - \mu\sin\theta}\right)} \le
|v| \le
\sqrt{rg\left(\frac{\sin\theta - \mu\cos\theta}{\cos\theta + \mu\sin\theta}\right)}
}\end{center}


or, dividing out $\cos\theta$ to involve $\tan\theta$ only:
\begin{center}\boxed{\sqrt{rg\left(\frac{\tan\theta + \mu}{1 - \mu\tan\theta}\right)} \le
|v| \le
\sqrt{rg\left(\frac{\tan\theta - \mu}{1 + \mu\tan\theta}\right)}
}\end{center}


\section{New Business -- Lightnight McQuation}


\end{document}